---
title: "RStudio"
author: "Marc A.T. Teunis"
date: "`r Sys.Date()`"
output:
  ioslides_presentation: 
    highlight: pygments
    widescreen: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## RStudio IDE
RStudio is an Integrated Development Environment that helps you be more productive with R. To use RStudio you need to first install R itself:

The lastest version of R can be installed from the Comprehensive R Archive Network `CRAN`: 

https://cran.r-project.org/

## Tutorials
Several good tutorials on starting with RStudio:

 - https://www.sitepoint.com/introduction-r-rstudio/\
 

## Preview releases (daily builds)
For the brave and impatient:
RStudio releases it's daily builds (test versions) on:

https://www.rstudio.com/products/rstudio/download/preview/

Here you can review the latest changes and improvements of the IDE.

## Cheatsheet:
This link will download the pdf of the RStudio cheatsheet:
https://github.com/rstudio/cheatsheets/raw/master/rstudio-ide.pdf

## More important cheatsheets that can help:
https://www.rstudio.com/resources/cheatsheets/

## A few tricks 
Go over them to learn about a few options in RStudio

## If you are really stuck
If R hangs, restart R by: `Session` -> `Restart R`

If login into the RStudio Server fails, sent me an email, I will reset your account for you

## New File
You can create new files under `File`. Choose from the list the file type that you want to create.

## R-scripts
 - R-script; used to run single actions, functions, cleaning, file-managment, a single graph or automated actions. 
 - Use `source("path_to_r_script") to call an individual script. 
 - Runs begin-to-end, can contain dependencies on other files
 
## R Markdown

 - In lab 4 we saw tha value of RMarkdown
 - Exercises have to be made in RMarkdown during the course. 
 - It is a very flexible and multi-pupose format to combine code, text and code-output in one product
 - Suitable to generate different formats: MS Word, MS Power Point, HTML Slide shows, PDF, HTML documents, Websites, Blogs, Books
 - Able to include different types of code chunks: R, Python, Bash, Stan, Perl etc.
 - Out of scope of this course
 
## Clear Console
To clear the contents of the console: `Edit` -> `Clear Console`

## Clear the `Global Environment`
To clear the global environment you have two options:
Either by using a line of code (be carefull not to put this in code you share with other, you will accentially clear **THEIR** environment, which may cause substraction of Karma points)

```{r, eval=FALSE}
## create dummies:
x <- runif(n = 10)
y = rnorm(n = 100)
z = "this is character dummy"
## remove them
rm(list=ls())
```

Or in RStudio: Click the small `stick-broom` in the `Environment` pane

## The `stick-broom` icon 
You will find the `stick-broom` in other panes in RStudio as well, it clears the object in the `Plots` and the `Viewer` panes respectively.

## Background theme
To change the theme: `Tools` -> `Global Options` -> `Appearance` -> Choos an `Editor Theme` that is appealing to you. Dark background is most relaxing for your eyes.

## Inline output disable
When you RStudio for the fist time amnd you start a new RMarkdown, the code output is shown inline of the RMarkdown document in the cript editor. It is best to disable this: 
`Tools` -> `Global Options` -> `R Markdown` -> Uncheck the option "Show output inline for all ...."

## Reformat code
To force you to adhere to the code style formatting:
You can reformat code according the code style suggested by Hadley Wickham here:

using `Edit` -> `Reformat Code`
See what happens to the code formatting if you select the code below and apply reformatting to it.

## Linux (Bash or PowerShell in Windows) Terminal
Sometimes we want to able to do some operations in Linux. For example seeing the head of a large file, without loading it into R first, or install addional software (in Linux you would need root acces or so-called `sudo` privileges for this).

To start a new `Terminal`: `Tools` -> `Terminal` -> `New Terminal`
Try it below and copy (Ctrl + C), paste (Shift + Insert) the following Bash code in your new `Terminal`

```{Bash, eval = FALSE}
head -30 data/allijn.tsv
```

## Reading data - help
To read data RStudio has an interactive help: 
Choose `File` -> `Import Dataset` -> `From Text (readr)` or the appropriate file type

 - Choose the file that you want to import under the `Browse` button
 - An interactive Console will appear, choose the right settings such as headers on/off, delimiter etc. 
 - Much like in the import function from Excel you will see a preview of the datatable.  - In the bottom right corner the appropriate R code is displayed. 
 - You can copy that code to the clipboard and paste it into a script.
 - Now execute the script and the datafile loads. 
 - If you want you can first adapt the variable name that is chosen to store the datframe in

## Try load data
Try to load the datafile "/data/kiemingen.xlsx" the way described above.

Think about why you should copy - paste the R-code to a script to load a file!?



