---
title: "Several R plot examples"
author: "Marc A.T. Teunis"
date: "`r Sys.Date()`"
output:
  ioslides_presentation: 
    highlight: pygments
    widescreen: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      error = FALSE,
                      message = FALSE)

```

## R packages
```{r, echo=TRUE}
## CRAN
library(stats)
library(circlize) 
library(pheatmap)
library(ggplot2)
library(drc)
library(tidyverse)

## Bionconductor
## Example: BiocManager::install('DESeq2')
library(DESeq2)
library(GEOquery)
library(tximport)
library(tximportData)
library(pasilla)
library(ReactomePA)
# biocLite("vsn")
#install.packages("ashr")
```

## Circos plot
```{r}
source(here::here("code", "circos_example.R"))
```

## Heatmaps
```{r}
library(gplots)
library(geneplotter)

# Mol Biol Cell. 1998 Dec;9(12):3273-97.
# Comprehensive identification of cell cycle-regulated genes of the yeast
# Saccharomyces cerevisiae by microarray hybridization.
# Spellman PT, Sherlock G, Zhang MQ, Iyer VR, Anders K, Eisen MB, Brown PO, Botstein D, Futcher B.
# Department of Genetics, Stanford University Medical Center, Stanford, California 94306-5120, USA.
#
# We sought to create a comprehensive catalog of yeast genes whose transcript levels vary
# periodically within the cell cycle. To this end, we used DNA microarrays and samples from
# yeast cultures synchronized by three independent methods: alpha factor arrest, elutriation,
# and arrest of a cdc15 temperature-sensitive mutant. Using periodicity and correlation algorithms,
# we identified 800 genes that meet an objective minimum criterion for cell cycle regulation.
# In separate experiments, designed to examine the effects of inducing either the G1 cyclin
# Cln3p or the B-type cyclin Clb2p, we found that the mRNA levels of more than half of these
# 800 genes respond to one or both of these cyclins. Furthermore, we analyzed our set of cell
# cycle-regulated genes for known and new promoter elements and show that several known elements
# (or variations thereof) contain information predictive of cell cycle regulation. A full
# description and complete data sets are available at http://cellcycle-www.stanford.edu

# read the data in from URL
bots <- read.table(url("http://genome-www.stanford.edu/cellcycle/data/rawdata/combined.txt"), sep="\t", header=TRUE)

# get just the alpha data
abot <- bots[,c(8:25)]
rownames(abot) <- bots[,1]
abot[1:7,] %>% as_tibble()

# get rid of NAs
abot[is.na(abot)] <- 0

# we need to find a way of reducing the data. Can't do ANOVA as there are no
# replicates. Sort on max difference and take first 1000
min <-apply(abot, 1, min)
max <- apply(abot, 1, max)
sabot <- abot[order(max - min, decreasing=TRUE),][1:1000,]

# cluster on correlation
cdist <- as.dist(1 - cor(t(sabot)))
hc <- hclust(cdist, "average")
```

## Draw a heatmap
```{r}
# x11()
heatmap.2(as.matrix(sabot),
          Rowv=as.dendrogram(hc),
          Colv=FALSE,
          cexRow=0.05,
          cexCol=1,
          dendrogram="row",
          scale="row",
          trace="none",
          density.info="none",
          key=FALSE,
          col=greenred.colors(80))
```

## Different colours
```{r}
# x11()
heatmap.2(as.matrix(sabot),
          Rowv=as.dendrogram(hc),
          Colv=FALSE,
          cexRow=0.05,
          cexCol=1,
          dendrogram="row",
          scale="row",
          trace="none",
          density.info="none",
          key=FALSE,
          col=heat.colors(80))
```

## More control over colours
```{r}
# x11()

my_palette <- colorRampPalette(c("blue", "yellow", "darkgreen"))(n = 80)

heatmap.2(as.matrix(sabot),
          Rowv=as.dendrogram(hc),
          Colv=FALSE,
          cexRow=0.05,
          cexCol=1,
          dendrogram="row",
          scale="row",
          trace="none",
          density.info="none",
          key=FALSE,
          col=my_palette)
```

## Dose response relations
```{r, echo=TRUE}
library(drc)
ryegrass %>% as_tibble()
## Fitting models to be plotted below
ryegrass.m1 <- drm(rootl~conc, data = ryegrass, fct = LL.4())
ryegrass.m2 <- drm(rootl~conc, data = ryegrass, fct = LL.3())  
# lower limit fixed at 0
```

## Plotting observations and fitted curve for the first model
```{r}
library(drc)
plot(ryegrass.m1, broken = TRUE)
```

## Adding fitted curve for the second model (not much difference)
```{r}
plot(ryegrass.m1, broken = TRUE)
plot(ryegrass.m2, broken = TRUE, add = TRUE, type = "none", col = 2, lty = 2)
```

## Add confidence region for the first model.
```{r}
plot(ryegrass.m1, broken = TRUE)
plot(ryegrass.m2, broken = TRUE, add = TRUE, type = "none", col = 2, lty = 2)
plot(ryegrass.m1, broken = TRUE, type="confidence", add=TRUE) 
```

## Finetuning the axis break
```{r}
plot(ryegrass.m1, broken = TRUE, bcontrol = list(style = "gap"))
```

## Pathway analysis
```{r}
library(ReactomePA)
## try http:// if https:// URLs are not supported
source("https://bioconductor.org/biocLite.R")
#biocLite("ReactomePA")
#biocLite("DOSE")
#biocLite("org.Hs.eg.db")
#browseVignettes("ReactomePA")

library(org.Hs.eg.db)
library(DOSE)
library(ReactomePA)

## ------------------------------------------------------------------------
library(ReactomePA)
data(geneList)
de <- names(geneList)[abs(geneList) > 1.5]
head(de)
x <- enrichPathway(gene=de,pvalueCutoff=0.05, readable=T)
head(as.data.frame(x))
```

## Barplot
```{r, fig.height=6, fig.width=12}
## ----fig.height=6, fig.width=12------------------------------------------

barplot(x, showCategory=8)

```

## Network with clusters
```{r}
## ----fig.height=10, fig.width=10-----------------------------------------
emapplot(x)
```

## Cnet network
```{r}
## ----fig.height=8, fig.width=8-------------------------------------------
cnetplot(x, categorySize="pvalue", foldChange=geneList)
```

## ggplot2
```{r}
library(ggplot2)
library(dplyr)
library(gapminder)
library(wesanderson)

gap_07 <- filter(gapminder, year == 2007)

ggplot(gap_07, aes(x = gdpPercap, y = lifeExp,
                   color = continent)) +

  geom_point(size = 2) +
  scale_x_log10() +
  theme_light() +
  theme(legend.position = c(0.1, 0.85),
legend.key = element_blank(),
axis.text = element_text(size = 12),
axis.title = element_text(size = 14)) +
  labs(x = "Per capita GDP",
       y = "Life Expectancy",
       title = "2007 Life Expectancy and GDP",
        color = "Continent") +
  ggsave("life_exp_gdp_2007_cutom.png", 
width = 7, height = 7)
```

## Time series
```{r, echo=FALSE}
#gapminder %>% names
gapminder %>% 
  group_by(continent, year) %>%
  summarise(total_pop_continent = sum(as.numeric(pop), na.rm = TRUE)) %>%
  ggplot(aes(x = year,
             y = total_pop_continent)) +
  geom_point(aes(colour = continent)) +
  geom_line(aes(group = continent, colour = continent)) +
  ylab("Total population per continent") +
  theme_bw()
```


## Wells plates
```{r}
# install.packages("platetools")
library(platetools)
library(viridis)

# example dataframe
df <- data.frame(vals = rnorm(384),
                 well = num_to_well(1:384, plate = 384))

raw_map(data = df$vals,
        well = df$well,
        plate = 384) +
    ggtitle("Example 384-well plate") +
    theme_dark() +
    scale_fill_viridis()
```

```{r, include=FALSE}
library(seqinr)
## Workflow to create senquence logo's in R
# https://davetang.org/muse/2013/01/30/sequence-logos-with-r/ 
library(Biostrings)  
library(seqLogo)

# Create position weight matrix

## Example sequences
sequences <- c("GCCAAAAAAA",
               "CGCAAAAAAA",
               "TCATAAAAAC",
               "GCATTAAAAA",
               "TCGTAAAAAC")
               
#?consensusMatrix

sequences <- DNAStringSet(sequences)
cmx <- consensusMatrix(sequences)
cmx

hasOnlyBaseLetters(sequences)

clean_cmx <- alphabetFrequency(sequences, baseOnly=TRUE)
               
#install if necessary
#source("http://bioconductor.org/biocLite.R")
#biocLite("seqLogo")



#create data frame using the four vectors

#define function that divides the frequency by the row sum i.e. proportions
proportion <- function(x){
  rs <- sum(x);
  return(x / rs);
}



#create position weight matrix
clean_cmx <- t(clean_cmx)
clean_cmx <- as.data.frame(clean_cmx[c(1:4),])
clean_cmx

pwm <- apply(clean_cmx, 2, proportion)
pwm
pwm <- makePWM(pwm)
```

## Map
```{r, echo=FALSE}
library(tmap)
library(tmaptools)

data(World, metro, rivers, land)

tmap_mode("plot")
## tmap mode set to plotting
tm_shape(land) +
    tm_raster("elevation", palette = terrain.colors(10)) +
tm_shape(World) +
    tm_borders("white", lwd = .5) +
    tm_text("iso_a3", size = "AREA") +
tm_shape(metro) +
    tm_symbols(col = "red", size = "pop2020", scale = .5) +
tm_legend(show = FALSE)

```

## Sequence logo plot
```{r}
#png("hunchback.png")
seqLogo(pwm)
# dev.off()

```

