---
title: "Rules for programming"
author: "Marc A.T. Teunis"
date: "`r Sys.Date()`"
output:
  ioslides_presentation: 
    highlight: pygments
    widescreen: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 5, fig.height = 3)
```

```{r, root_1, include=FALSE}
## defines the root of the project for later use
require("rprojroot") || utils::install.packages("rprojroot")
library(rprojroot)
root <- find_root_file(criterion = is_rstudio_project)
```

## Simple rules to start with

 * Write code for humans, NOT for computers!
 * Keep everything in the same place (work with projects)
 * Do not repeat yourself (more than two times --> write a function)
 * Do on thing at a time
 * Take a break if it get's too hairy

## Guerilla analytics principles
Specific guidelines from Ridge

* Use a simple, relatively flat folder lay-out for the project
* Never edit raw data files received from the 'client'
* Use a data log to track metadata of received data sets
* Store supporting material near the data 
* Add row IDs to source data when merging datasets
* Keep original file names
* Load data as plain text and typecast as little as possible (all lower case)
* Do everything programmatically
* Keep original columns and add edited versions of a column as a new column

## Project folder structure

 * Ridge promotes the use of a clear and simple project folder structure. Its
structure should be simple to use and transparant for all members of the
analytics team. 

 * The data manipulation environment (DME) is the area where all project data is manipulated and analysed. It can be a formal database management system or a folder in the file system. 

 * The DME for this course is the **Global Environment** of RStudio Server

 * The raw data is stored in the **data-raw** folder. Edited data or genearated datasets (cleaning etc.) are stored in the **data** folder.
 
## A project folder example
| Folder       | Description  |
|--------------|----------------------------------------------------------------------------------------------|
| **/R**   | The project code used for loading and analyzing data. |
| **/inst** | Generic code of utilities and functions used throughout the project. |
| **/data-raw** | Raw data as received. Note that data sets are stored in a folder with their unique ID (e.g. D010). Also note that each data set folder contains a folder **/data-raw/supporting** that contains supporting material. The metadata of the data sets is maintained in a data log (**data-log.xlsx**). |
| **/data** | A folder contained edited and generated data |
| **/doc** | The documents that result from overall analyses (e.g. like an R markdown document). |
| **/Vignettes** | A folder similar in structure to **data-raw** with the different work products that are produced within the project. A work product is a single unit of analysis with a clearly defined product to be delivered. The metadata of the work products is maintained in a work products log (**work-products-log.xlsx**). |

## R package
This structure follows the R package structure

To illustrate this we will start a new package:

 1. Think about a good name (max 8 letter, lowercase) 
 2. Initiate a new R package: File -> New Project -> Package using Devtools
 3. Look at the foldernames in the new package Project
 4. This is a standard folder structure that provides a clear framework for managing files 
 5. Adhere to this folder-structure, even when not working in a package.
 
 





